---
title: "2_Exploring Longitudinal Data on Change"
author: "Amol Jadhav"
date: "December 23, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import data
```{r}
tolerance <- read.csv(file.choose(), header=TRUE)
head(tolerance)

# Inputting and printing the tolerance "person-level" data
tolerance.pp <- read.csv(file.choose(), header=TRUE)

attach(tolerance.pp)
print(tolerance.pp[c(1:9, 76:80), ])
```

# Analysis
```{r}
#Bivariate correlation among tolerance scores assessed on five occassions
cor(tolerance[ , 2:6])

# Emprirical growth plots
library(lattice) # load lattice library for xyplot function

xyplot(tolerance ~ age | id, data=tolerance.pp, as.table=T)

# Smooth nonparametric trajectories superimposed on empirical growth plots
xyplot(tolerance~age | id, data=tolerance.pp,
      prepanel = function(x, y) prepanel.loess(x, y, family="gaussian"),
      xlab = "Age", ylab = "Tolerance",
      panel = function(x, y) {
      panel.xyplot(x, y)
      panel.loess(x,y, family="gaussian") },
      ylim=c(0, 4), as.table=T)
```
# Fitting separate within person OLS regression models
```{r}
by(tolerance.pp, id, function(x) summary(lm(tolerance ~ time, data=x)))
```

# Plots
```{r}
# Plot for the intercepts
int <- by(tolerance.pp, id, function(data)
                  coefficients(lm(tolerance ~ time, data = data))[[1]])
int <- unlist(int)
names(int) <- NULL
summary(int)

plot(int)

# Plot for fitted rate of change
rate <- by(tolerance.pp, id, function(data)
              coefficients(lm(tolerance ~ time, data = data))[[2]])
rate <- unlist(rate)
names(rate) <- NULL
summary(rate)

plot(rate)

# Plot for R sq
rsq <- by(tolerance.pp, id, function(data)
             summary(lm(tolerance ~ time, data = data))$r.squared)
rsq <- unlist(rsq)
names(rsq) <- NULL
summary(rsq)

barplot(rsq)
```

# Fitted OLS trajectories superimposed on the empirical growth plots
```{r}
attach(tolerance)

xyplot(tolerance ~ age | id, data=tolerance.pp, 
  panel = function(x, y){
    panel.xyplot(x, y)
    panel.lmline(x, y)
  }, ylim=c(0, 4), as.table=T)
```

# The collection of the trajectories of the raw data in the top panel; in the bottom panel is the collection of fitted OLS trajectories
```{r}
#plot of the raw data

interaction.plot(tolerance.pp$age, tolerance.pp$id, tolerance.pp$tolerance) 

# fitting the linear model by id
fit <- by(tolerance.pp, id, function(bydata) fitted.values(lm(tolerance ~ time, data=bydata))) 
fit <- unlist(fit)

# plotting the linear fit by id
interaction.plot(age, id, fit, xlab="age", ylab="tolerance")

```

# Descriptive statistics of the estimates obtained by fitting the linear model by id
```{r}
#obtaining the intercepts from linear model by id
ints <- by(tolerance.pp, tolerance.pp$id, 
           function(data) coefficients(lm(tolerance ~ time, data=data))[[1]])  
ints1 <- unlist(ints)
names(ints1) <- NULL
mean(ints1)  # 1.3577

sqrt(var(ints1)) # 0.29777


#obtaining the slopes from linear model by id
slopes <- by(tolerance.pp, tolerance.pp$id,
             function(data) coefficients(lm(tolerance ~ time, data=data))[[2]])  
slopes1 <- unlist(slopes)
names(slopes1) <- NULL
mean(slopes1)  # 0.13081

```

# OLS fitted trajectories separated by levels of selected predictors. The two first panels are separated by gender; the two last are separated by levels of exposure.
```{r}
# fitting the linear model by id, males only
tolm <- tolerance.pp[male==0 , ]
fitmlist <- by(tolm, tolm$id, function(bydata) fitted.values(lm(tolerance ~ time, data=bydata))) 
fitm <- unlist(fitmlist)

#appending the average for the whole group
lm.m <- fitted( lm(tolerance ~ time, data=tolm) )
names(lm.m) <- NULL
fit.m2 <- c(fitm, lm.m[1:5])
age.m <- c(tolm$age, seq(11,15))
id.m <- c(tolm$id, rep(111, 5))

#plotting the linear fit by id, males
#id.m=111 denotes the average value for males
interaction.plot(age.m, id.m, fit.m2, ylim=c(0, 4), xlab="AGE", ylab="TOLERANCE", lwd=1)
title(main="Males")


#fitting the linear model by id, females only
tol.pp.fm <- tolerance.pp[tolerance.pp$male==1 , ]
fit.fm <- by(tol.pp.fm, tol.pp.fm$id, 
          function(data) fitted.values(lm(tolerance ~ time, data=data))) 
fit.fm1 <- unlist(fit.fm)
names(fit.fm1) <- NULL

#appending the average for the whole group
lm.fm <- fitted( lm(tolerance ~ time, data=tol.pp.fm) )
names(lm.fm) <- NULL
fit.fm2 <- c(fit.fm1, lm.fm[1:5])
age.fm <- c(tol.pp.fm$age, seq(11,15))
id.fm <- c(tol.pp.fm$id, rep(1111, 5))

#plotting the linear fit by id, females
#id.fm=1111 denotes the average for females
interaction.plot(age.fm, id.fm, fit.fm2, ylim=c(0, 4), xlab="AGE", ylab="TOLERANCE", lwd=1)
title(main="Females")
```

```{r}
#fitting the linear model by id, low exposure
tol.pp.low <- tolerance.pp[tolerance.pp$exposure < 1.145 , ]
fit.low <- by(tol.pp.low, tol.pp.low$id, 
          function(data) fitted.values(lm(tolerance ~ time, data=data))) 
fit.low1 <- unlist(fit.low)
names(fit.low1) <- NULL

#appending the average for the whole group
lm.low <- fitted( lm(tolerance ~ time, data=tol.pp.low) )
names(lm.low) <- NULL
fit.low2 <- c(fit.low1, lm.low[1:5])
age.low <- c(tol.pp.low$age, seq(11,15))
id.low <- c(tol.pp.low$id, rep(1, 5))

#plotting the linear fit by id, low Exposure
#id.low=1 denotes the average for group
interaction.plot(age.low, id.low, fit.low2, ylim=c(0, 4), xlab="AGE", ylab="TOLERANCE", lwd=1)
title(main="Low Exposure")


#fitting the linear model by id, high exposure
tol.pp.hi <- tolerance.pp[tolerance.pp$exposure >= 1.145 , ]
fit.hi <- by(tol.pp.hi, tol.pp.hi$id, 
          function(data) fitted.values(lm(tolerance ~ time, data=data))) 
fit.hi1 <- unlist(fit.hi)
names(fit.hi1) <- NULL

#appending the average for the whole group
lm.hi <- fitted( lm(tolerance ~ time, data=tol.pp.hi) )
names(lm.hi) <- NULL
fit.hi2 <- c(fit.hi1, lm.hi[1:5])
age.hi <- c(tol.pp.hi$age, seq(11,15))
id.hi <- c(tol.pp.hi$id, rep(1, 5))

#plotting the linear fit by id, High Exposure
#id.hi=1 denotes the average for group
interaction.plot(age.hi, id.hi, fit.hi2, ylim=c(0, 4), xlab="AGE", ylab="TOLERANCE", lwd=1)
title(main="High Exposure")

```

# OLS estimates plotted against the predictors male and exposure
```{r}
#Using the slopes and intercepts from the linear model fitted by id 
plot(tolerance$male, ints1, xlab="Male", ylab="Fitted initial status", 
     xlim=c(0, 1), ylim=c(0.5, 2.5))

cor(tolerance$male, ints1) #0.00863

plot(tolerance$exposure, ints1, xlab="Exposure", ylab="Fitted initial status", 
     xlim=c(0, 2.5), ylim=c(0.5, 2.5))

cor(tolerance$exposure, ints1) #0.23244

plot(tolerance$male, slopes1, xlab="Male", ylab="Fitted rate of change", 
     xlim=c(0, 1), ylim=c(-0.4, .8))

cor(tolerance$male, slopes1) #0.19357

plot(tolerance$exposure, slopes1, xlab = "Exposure", ylab = 
	"Fitted rate of change", xlim = c(0, 2.5), ylim = c(-0.2, 0.8))

cor(tolerance$exposure, slopes1) #0.4421
```










