---
title: "4_Data analysis multilevel model for change"
author: "Amol Jadhav"
date: "December 23, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# import Data
```{r}
alcohol1 <- read.table(file.choose(), header=T, sep=",")
attach(alcohol1)
``

#Empirical growth plots with superimposed OLS trajectories
```{r}

library(lattice)

xyplot(alcuse~age | id, 
   data=alcohol1[alcohol1$id %in% c(4, 14, 23, 32, 41, 56, 65, 82), ], 
   panel=function(x,y){
	panel.xyplot(x, y)
	panel.lmline(x,y)
}, ylim=c(-1, 4), as.table=T)

```

# Fitted OLS trajectories displayed separately by coa status and peer levels. Upper left panel, coa=0
```{r}
alcohol.coa0 <- alcohol1[alcohol1$coa==0, ]

#fitting the linear model by id
f.coa0 <- by(alcohol.coa0, alcohol.coa0$id, 
                function(data) fitted(lm(alcuse~age, data=data)))
#transforming f.coa from a list to a vector and 
#stripping of the names of the elements in the vector
f.coa0 <- unlist(f.coa0)
names(f.coa0) <- NULL

#plotting the linear fit by id
interaction.plot(alcohol.coa0$age, alcohol.coa0$id, f.coa0, 
                 xlab="AGE", ylab="ALCUSE", ylim=c(-1, 4), lwd=1)
title("COA=0")

#2
alcohol.coa1 <- alcohol1[alcohol1$coa==1, ]

#fitting the linear model by id
f.coa1 <- by(alcohol.coa1, alcohol.coa1$id, 
             function(data) fitted(lm(alcuse~age, data=data)))
#transforming f.coa1 from a list to a vector and 
#stripping of the names of the elements in the vector
f.coa1 <- unlist(f.coa1)
names(f.coa1) <- NULL

#plotting the linear fit by id
interaction.plot(alcohol.coa1$age, alcohol.coa1$id, f.coa1, 
                 xlab="AGE", ylab="ALCUSE", ylim=c(-1, 4), lwd=1)
title("COA=1")
mean(alcohol1$peer) #1.0175

#3
alcohol.lowpeer <- alcohol1[alcohol1$peer<=1.01756, ]

#fitting the linear model by id
f.lowpeer <- by(alcohol.lowpeer, alcohol.lowpeer$id, 
             function(data) fitted(lm(alcuse~age, data=data)))
#transforming f.lowpeer from a list to a vector and 
#stripping of the names of the elements in the vector
f.lowpeer <- unlist(f.lowpeer)
names(f.lowpeer) <- NULL

#plotting the linear fit by id
interaction.plot(alcohol.lowpeer$age, alcohol.lowpeer$id, f.lowpeer, 
                 xlab="AGE", ylab="ALCUSE", ylim=c(-1, 4), lwd=1)
title("Low Peer")

#4
alcohol.hipeer <- alcohol1[alcohol1$peer>1.01756, ]

#fitting the linear model by id
f.hipeer <- by(alcohol.hipeer, alcohol.hipeer$id, 
               function(data) fitted(lm(alcuse~age, data=data)))
#transforming f.hipeer from a list to a vector and 
#stripping of the names of the elements in the vector
f.hipeer <- unlist(f.hipeer)
names(f.hipeer) <- NULL

#plotting the linear fit by id
interaction.plot(alcohol.hipeer$age, alcohol.hipeer$id, f.hipeer, 
                 xlab="AGE", ylab="ALCUSE", ylim=c(-1, 4), lwd=1)
title("High Peer")

```


# Models
```{r}
#Model A
library(nlme)

model.a <- lme(alcuse~ 1, alcohol1, random= ~1 |id)
summary(model.a)

#Model B
model.b <- lme(alcuse ~ age_14 , data=alcohol1, random= ~ age_14 | id, method="ML")
summary(model.b)

#Model C
model.c <- lme(alcuse ~ coa*age_14 , data=alcohol1, random= ~ age_14 | id, method="ML")
summary(model.c)

#Model D
model.d <- lme(alcuse ~ coa*age_14+peer*age_14 , data=alcohol1, random= ~ age_14 | id, method="ML")
summary(model.d)

#Model E
model.e <- lme(alcuse ~ coa+peer*age_14 , data=alcohol1, random= ~ age_14 | id, method="ML")
summary(model.e)

#Model F
model.f <- lme(alcuse ~ coa+cpeer*age_14 , data=alcohol1, random= ~ age_14 | id, method="ML")
summary(model.f)

#Model G
model.g <- lme(alcuse ~ ccoa+cpeer*age_14 , data=alcohol1, random= ~ age_14 | id, method="ML")
summary(model.g)
```

# Plots
```{r}
# Unconditional growth model
fixef.b <- fixef(model.b)
fit.b <- fixef.b[[1]] + alcohol1$age_14[1:3]*fixef.b[[2]]
plot(alcohol1$age[1:3], fit.b, ylim=c(0, 2), type="b", 
     ylab="predicted alcuse", xlab="age")   
title("Model B \n Unconditional growth model")

# uncontrolled effects
fixef.c <- fixef(model.c)
fit.c0 <- fixef.c[[1]] + alcohol1$age_14[1:3]*fixef.c[[3]]
fit.c1 <- fixef.c[[1]] + fixef.c[[2]] + 
          alcohol1$age_14[1:3]*fixef.c[[3]] +
          alcohol1$age_14[1:3]*fixef.c[[4]]
plot(alcohol1$age[1:3], fit.c0, ylim=c(0, 2), type="b", 
     ylab="predicted alcuse", xlab="age")
lines(alcohol1$age[1:3], fit.c1, type="b", pch=17)   
title("Model C \n Uncontrolled effects of COA") 
legend(14, 2, c("COA=0", "COA=1")

# Model for the controlled effects of COA
library(nlme)
fixef.e <- fixef(model.e)
fit.ec0p0 <- fixef.e[[1]] + .655*fixef.e[[3]] +
             alcohol1$age_14[1:3]*fixef.e[[4]] +
             .655*alcohol1$age_14[1:3]*fixef.e[[5]]   
fit.ec0p1 <- fixef.e[[1]] + 1.381*fixef.e[[3]] +
             alcohol1$age_14[1:3]*fixef.e[[4]] +
             1.381*alcohol1$age_14[1:3]*fixef.e[[5]] 
fit.ec1p0 <- fixef.e[[1]] + fixef.e[[2]] + .655*fixef.e[[3]] +
             alcohol1$age_14[1:3]*fixef.e[[4]] +
             .655*alcohol1$age_14[1:3]*fixef.e[[5]] 
fit.ec1p1 <- fixef.e[[1]] + fixef.e[[2]] + 1.381*fixef.e[[3]] +
             alcohol1$age_14[1:3]*fixef.e[[4]] +
             1.381*alcohol1$age_14[1:3]*fixef.e[[5]]
plot(alcohol1$age[1:3], fit.ec0p0, ylim=c(0, 2), type="b", 
     ylab="predicted alcuse", xlab="age", pch=2)
lines(alcohol1$age[1:3], fit.ec0p1, type="b", pch=0)   
lines(alcohol1$age[1:3], fit.ec1p0, type="b", pch=17)   
lines(alcohol1$age[1:3], fit.ec1p1, type="b", pch=15)   
title("Model E \n *Final* model for the controlled effects of COA") 
legend(14, 2, c("COA=0, low peer", "COA=0, high peer", 
       "COA=1, low peer", "COA=1, high peer"))

```

# Model analysis
```{r}
#creating the residuals (epsilon.hat)
resid <- residuals(model.f)
qqnorm(resid)

#creating the standardized residual (std epsilon.hat)
resid.std <- resid/sd(resid)
plot(alcohol1$id, resid.std, ylim=c(-3, 3), ylab="std epsilon hat")
abline(h=0)

#extracting the random effects of model f
ran <- random.effects(model.f)
qqnorm(ran[[1]])

#standardizing the ksi0i.hat
ran1.std <- ran[[1]]/sd(ran[[1]])
plot(alcohol1$id[alcohol1$age==14], ran1.std, ylim=c(-3, 3), ylab="std psi_0i hat")
abline(h=0)

qqnorm(ran[[2]])

#standardizing the ksi1i.hat
ran2.std <- ran[[2]]/sd(ran[[2]])
plot(id[alcohol1$age==14], ran2.std, ylim=c(-3, 3), ylab="std psi_1i hat")
abline(h=0)

plot(alcohol1$age, resid, ylim=c(-2, 2), ylab="epsilon.hat", 
     xlab="AGE")
abline(h=0)

plot(alcohol1$coa[alcohol1$age==14], ran[[1]], ylim=c(-1, 1), 
     ylab="ksi0i.hat", xlab="COA")
abline(h=0)

plot(alcohol1$peer[alcohol1$age==14], ran[[1]], ylim=c(-1, 1), 
     xlim=c(0, 3), ylab="ksi0i.hat", xlab="PEER")
abline(h=0)


plot(alcohol1$coa[alcohol1$age==14], ran[[2]], ylim=c(-1, 1), 
     ylab="ksi1i.hat", xlab="COA")
abline(h=0)

plot(alcohol1$peer[alcohol1$age==14], ran[[2]], ylim=c(-1, 1), 
     xlim=c(0, 3), ylab="ksi1i.hat", xlab="PEER")
abline(h=0)
```

